---
title: "Post-Midterm Review: Building Blocks of Modern Regression"
subtitle: "Modules 1–3: LMMs → GLM(M)s → Marginal Models"
format:
  revealjs:
    theme: simple
    slide-number: true
    incremental: true
    smaller: true
execute: 
  echo: true
  warning: false
---

## Session Goals

-   Diagnose and clarify conceptual misunderstandings from the midterm\
-   Reconnect theory to model structure\
-   Distinguish **hierarchical**, **generalized**, and **marginal** perspectives\
-   Prepare for nonlinear extensions (splines, GAMs)

::: callout-tip
By the end of class, you should be able to:\
1. Identify the mean, variance, and correlation structure of any model\
2. Explain what parameter each model estimates (subject vs. population)\
3. Interpret coefficients on the correct scale
:::

------------------------------------------------------------------------

## Warm-Up Poll

### Quick Concept Check

1.  What changes when we move from LM → LMM?\
2.  Does the link function act on $y$ or on $E[y|x]$?\
3.  Interpret $\text{Var}(y_{ij}) = \sigma_u^2 + \sigma_\epsilon^2$.\
4.  Why might GEE and GLMM yield different $\beta$'s?\
5.  When does GLS outperform WLS?

::: callout-note
**Instructor Prompt:** Discuss reasoning, not just correctness.\
Surface misconceptions before deep dives.
:::

------------------------------------------------------------------------

## Module 1 --- Linear Mixed Models {.smaller}

::: callout-important
**Motivation**

-   Longitudinal data
-   Clustered data
-   Normal Outcomes
:::

------------------------------------------------------------------------

## Model

\tiny

::: fragment
We model hierarchical/correlated data:\

$$
y_{ij} = \beta_0 + \beta_1 x_{ij} + \theta_i + \epsilon_{ij},
\quad \theta_i\sim N(0,\tau^2),\ 
\epsilon_{ij}\sim N(0,\sigma_\epsilon^2),\
\theta_i \perp \epsilon_{ij}
$$
:::

**Interpretation**

::: fragment
| Concept | Meaning |
|:-----------------------------------|:-----------------------------------|
| **Population-level effects** | $\beta_0, \beta_1$: overall (fixed) trend across all groups |
| **Random effects** | $\theta_i$: group-level deviations (random across groups) |
| **Overall trend** | $\beta_0 + \beta_1 x_{ij}$ |
| **Conditional, group-specific trend** | $(\beta_0 + \theta_i) + \beta_1 x_{ij}$ |
| **Total variability** | $\tau^2 + \sigma_\epsilon^2$ |
| **Within-group variability** | $\sigma_\epsilon^2$ |
:::

------------------------------------------------------------------------

## 

**Key Ideas** - Two error sources →

-   variability within clusters $\sigma^2$\
-   variability between clusters $\tau^2$\
-   shrinkage!

Questions for you?

-   What is the covariance between two different clusters?
-   What is the covariance between two units/subjects within the same cluster?
-   What is the overall variability of the outcome across the entire population?

**Intra-class correlation:** the proportion of total variability explained by group membership.

-   Correlation between two individuals in the same group: $$
    \rho = \frac{\tau2}{\tau^2+\sigma_\epsilon^2}
    $$

------------------------------------------------------------------------

## Shrinkage

**Best Linear Unbiased Predictor**: Derived through partial pooling.

-   What is the main concept behind pooling?

-   $\hat{\theta}_i = B_i \cdot \bar{y}_i + (1-B_i) \cdot \hat{\mu}$

-   The shrinkage factor is the **precision weight** $B_i$:

::: fragment
$$
B_i = \frac{n_i}{n_i + \sigma^2/\tau^2}
$$

-   If $B_i$ = large, this means large precision, and higher weight on the data.
-   If $B_i$ = small, this means smaller precision, and higher weight on the population mean.
:::

## 

```{r}
#| echo: true
#| results: 'markup'  # ensures pretty formatting
#| code-fold: true    # collapses the full output (optional)
library(lme4)
set.seed(1)
group <- factor(rep(1:6, each = 5))
x <- rnorm(30)
u <- rnorm(6, 0, 1)[group]
y <- 5 + 2*x + u + rnorm(30, 0, 1)
m1 <- lmer(y ~ x + (1|group))
summary(m1)$coefficients   # print only fixed effects
VarCorr(m1)    
```

::: callout-tip
Visualize shrinkage:\
`coef(m1)$group` vs. group means from `lm(y ~ x + group)`
:::

```{r}
#| echo: false
#| results: 'markup'  # ensures pretty formatting
#| code-fold: true    # collapses the full output (optional)
library(lme4)

set.seed(1)
group <- factor(rep(1:6, each = 5))
x <- rnorm(30)
u <- rnorm(6, 0, 1)[group]
y <- 5 + 2*x + u + rnorm(30, 0, 1)

m1 <- lmer(y ~ x + (1|group))
lm_fit <- lm(y ~ x + group)

# Unpooled group means at x = 0, one per group (includes the reference)
newdat <- data.frame(x = 0, group = levels(group))
lm_means <- predict(lm_fit, newdata = newdat)

# Partially pooled random-intercept estimates (same order as levels(group))
re_means <- coef(m1)$group[,"(Intercept)"]

# Sanity check: same length and aligned by group
stopifnot(length(lm_means) == length(re_means))

plot(lm_means, re_means,
     xlab = "Unpooled group means (lm, at x=0)",
     ylab = "Partially pooled intercepts (lmer)",
     main = "Visualizing Shrinkage",
     pch = 19)
abline(0, 1, lty = 2)
text(lm_means, re_means, labels = levels(group), pos = 3, cex = 0.8)
 
```

------------------------------------------------------------------------

## Parameter Estimation

::::: fragment
-   How do we estimate the population-level parameters?

-   **Best Linear Unbiased Estimator of the Coefficient** $$
    \hat{\beta} = (X^T V^{-1} X)^{-1} X^{T} V^{-1}y,\ V \neq \sigma^2 I
    $$

-   **Maximum Likelihood Estimate** $\sigma^2$

::: fragment
$$
\hat{\sigma}^2 =\frac{1}{n} \sum_{i=1}^n (y_i - \bar{y})^2
$$
:::

-   **Restricted Maximum Likelihood Estimate** $\sigma^2$

::: fragment
$$
\hat{\sigma}_{REML}^2 =\frac{1}{n-1} \sum_{i=1}^n (y_i - \bar{y})^2
$$
:::
:::::

------------------------------------------------------------------------

## Module 2A--B --- Generalized Linear Models

::: callout-important
**Motivation**

-   Longitudinal data
-   Clustered data
-   Non-Normal Count Outcomes
:::

------------------------------------------------------------------------

## Model

Unified framework:

$$
g(\mu_i) = X_i\beta, \qquad E(Y_i|X_i) = g^{-1}(X_i \beta), \qquad
\text{Var}(Y_i|X_i) = V(g^{-1}(X_i \beta)) = \phi V(\mu_i)
$$

| Distribution | Link     | Variance     |
|--------------|----------|--------------|
| Normal       | identity | constant     |
| Poisson      | log      | $\mu$        |
| Binomial     | logit    | $\mu(1-\mu)$ |

- What are the properties of the link function?

------------------------------------------------------------------------


```{r, echo=F}
df <- data.frame(x = rnorm(100))
df$y <- rpois(100, lambda = exp(0.5 + 0.3*df$x))
df$N <- sample(x = seq(50, 100), size = 100, replace = TRUE)
```

```{r}
#| echo: True
#| results: 'markup'  # ensures pretty formatting
#| code-fold: true    # collapses the full output (optional)
# Poisson example
m_poiss <- glm(y ~ x, family = poisson(link="log"), offset = log(df$N), data=df)
summary(m_poiss)
```


```{r}
#| echo: True
#| results: 'markup'  # ensures pretty formatting
#| code-fold: true    # collapses the full output (optional)
# Binomial example
mbinom <- glm(cbind(y, N-y) ~ x, family = binomial(link="logit"), data=df)
summary(mbinom)
```


-----


## Interpretation

::: callout-note

\smaller
Interpret β₁: Both on original and transformed scales

- Each one-unit increase in *x* multiplies $E[y|x]$ by $\exp(β_1)$ 

- $\beta_1$ is interpreted as the change in _______-__________  per one unit change in *x*.

- Define an odds ratio

- Define an incidence rate ratio

- How do you get the 95\% CIs for the OR?
:::

------------------------------------------------------------------------


## Model Inference and Comparison {.smaller}

**Context:** In GLMs, we assess predictors and compare models using three key tools:

-  🔹 Wald Test: Tests **individual coefficients**:  
  $$
  z = \frac{\hat{\beta}_j}{\text{SE}(\hat{\beta}_j)} \quad \Rightarrow \quad p\text{-value from } N(0,1)
  $$
- Interpretation: is $\beta_j$ significantly different from 0?


-  🔹 Likelihood Ratio Test (LRT): Compares **nested models**:  
  $$
  G^2 = -2(\ell_{\text{reduced}} - \ell_{\text{full}}) \sim \chi^2_{df}
  $$
- Tests whether removing a term significantly worsens model fit.


-  🔹 Akaike Information Criterion (AIC): Compares **non-nested models** (general model selection):  $\text{AIC} = -2\ell + 2p$
- Balances model **fit** ($\ell$ = log-likelihood) and **parsimony** ($p$ = number of parameters).
- Lower AIC ⇒ better tradeoff between complexity and fit.

-----

**Summary Table**

| Criterion | Purpose | Uses | Test Statistic | Lower/Smaller = Better? |
|:--|:--|:--|:--|:--:|
| **Wald Test** | Test individual $\beta_j$ | Non-nested, single term | $z = \hat\beta / SE$ | – |
| **LRT** | Compare nested models | Nested | $G^2 = -2(\ell_0 - \ell_1)$ | ✔ |
| **AIC** | Compare any models | Nested / non-nested | $-2\ell + 2p$ | ✔ |


-----------------


## Module 2C--D GLMMs

Add random effects to handle within-cluster dependence: 
$$
g(E[y_{ij}|u_j]) = X_{ij}\beta + \theta_i,\quad \theta_i \sim N(0, \tau^2), \quad \text{NO } \epsilon!
$$

```{r}
#| echo: True
#| results: 'markup'  # ensures pretty formatting
#| code-fold: true    # collapses the full output (optional)
# Binary GLMM example
set.seed(2)
subject <- factor(rep(1:40, each = 5))
x <- rnorm(200)
u <- rnorm(40, 0, 0.75)[subject]
eta <- -1 + 1.2*x + u
y <- rbinom(200, 1, plogis(eta))
m_glmm <- glmer(y ~ x + (1|subject), family=binomial)
summary(m_glmm)
```

----------------- 

## GLM vs. GLMM: Population-Average vs. Subject-Specific {.smaller}

In a **GLM**, we model the *marginal* (population-average) relationship:

::::: fragment

$$
E[y_{ij} \mid x_{ij}] = g^{-1}(\beta_0 + \beta_1 x_{ij})
$$

:::

- The effect of $x$ (the slope $\beta_1$) reflects the **average association across all subjects**,  
because random subject differences are **integrated out** of the model.



In a **GLMM**, we model the *conditional* (group-specific) relationship:

::::: fragment

$$
E[y_{ij} \mid x_{ij}, \theta_i] = g^{-1}(\beta_0 + \beta_1 x_{ij} + \theta_i)
$$

:::

::: callout-tip
Compare $\hat{\beta}_x$ from: 

- glm(y ~ x) → Population-average (marginal)
- glmer(y ~ x + (1|subject)) → Subject-specific (conditional)

👉 Subject-specific (GLMM) effects are typically larger in magnitude,
because they remove between-subject variation and focus on within-subject change.
:::





------------------------------------------------------------------------

## Module 3 --- Marginal Models (WLS, GLS, GEE)


::: callout-important
**Motivation**

-   Longitudinal data
-   Clustered data
-   Want population-level associations only. But also want to account for correlation structure
:::

------------------------------------------------------------------------

### GEE Framework: Choosing a covariance structure

Progression of correlation handling:

| Method | Handles               | Target          |
|--------|-----------------------|-----------------|
| WLS    | heteroskedasticity: $V$ is diagonal | population mean |
| GLS    | known correlation   $V$ is any known positive def matrix  | population mean |
| GEE    | estimated correlation | population mean |

Correlation structure examples:  
- $V_i = \sigma^2 R_i(\alpha)$

    - Independent
    - Exchangeable
    - AR(1)
    - Unstructured


-----

## GLS Estimators

- Generalized Least Squares Estimator is the Best Linear Unbiased Estimator whe $V$ is correctly specified

::::: fragment

$$
\hat{\beta}_{GLS} = (X^T V^{-1}X)^{-1} X^T V^{-1}y
$$

$$
cov(\hat{\beta}_{GLS}) = (X^T V^{-1}X)^{-1} 
$$

:::

---

## What if our assumption about $V$ is wrong?


- If $V$ is **mis-specified**, $\hat{\beta}_{GLS}$ is still **consistent**,  
  but its **standard errors are wrong**.

- **GEE (Generalized Estimating Equations)** fixes this by using a
  **“working correlation”** and a **robust (sandwich) variance estimator.**

:::: {.fragment}
$$
\hat{V}_{robust} = (X^T V^{-1} X)^{-1} \left( \sum_i X_i^T V^{-1} \hat{\Omega}_iV^{-1} X_i \right) (X^T V^{-1} X)^{-1}
$$
where $\hat{\Omega}_i = \epsilon_i \epsilon_i^\top$ (cluster residual covariance)





::::


--------

## GEE Estimating Equations {.smaller}

The goal of GEE is to estimate **population-average effects** when observations within a cluster (e.g., repeated measures on a person) are **correlated**.

We solve for $\hat{\beta}$ that makes the following **weighted residuals sum to zero**:


$$
U(\beta) = \sum_{i=1}^m D_i^\top V_i^{-1} (y_i - \mu_i) = 0
$$


| Symbol | Meaning |
|:--|:--|
| $i$ | Cluster or subject index ($i = 1, \dots, m$) |
| $y_i$ | Vector of responses for cluster $i$ |
| $\mu_i = E(y_i)$ | Mean vector under the model |
| $D_i = \frac{\partial \mu_i}{\partial \beta}$ | Design (derivative) matrix |
| $V_i = A_i^{1/2} R_i(\alpha) A_i^{1/2}$ | Working covariance matrix of $y_i$ |
| $A_i$ | Diagonal matrix of marginal variances |
| $R_i(\alpha)$ | Working correlation (independent, exchangeable, AR(1), etc.) |

---

## Interpretation

- $U(\beta)$ generalizes the **score equation** from GLMs to correlated data.  
- Even if the **working correlation** $R_i(\alpha)$ is wrong:
- $\hat{\beta}$ is still **consistent** (mean model correct ⇒ unbiased population effect).  
- Only the **standard errors** are affected.  
  → Fix this with the **robust (sandwich) variance**.



------------

```{r}
library(geepack)
id <- factor(rep(1:30, each = 4))
time <- rep(1:4, 30)
x <- rnorm(120)
eta <- 0.5 + 0.4*x
y <- rpois(120, exp(eta))
gee_ind <- geeglm(y ~ x, id=id, family=poisson, corstr="independence")
gee_exc <- geeglm(y ~ x, id=id, family=poisson, corstr="exchangeable")
summary(gee_ind)$coef
summary(gee_exc)$coef
```

::: callout-note
β estimates are similar, but SEs differ by working correlation.\
GEE targets the **population-average** effect.
:::

------------------------------------------------------------------------

## Midterm Deconstruction & Takeaways


**Questions on the Midterm**

------------------------------------------------------------------------
